Dirace: Practical Proof Automation of Dirac Notation Equations

### 2025-01-25 main.tex
2025/1/25 GPT4 improved

%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}

    Dirac notation is the fundamental language in quantum computation and quantum information. Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate the equational reasoning with Dirac notation, a critical yet time-intensive task. This work bases upon and extends DiracDec, aiming to develop a solver optimized for practical applications. Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labelled Dirac notation. Our solver Dirace is implemented in \CC\ with a Mathematica backend, and demonstrates advantages in decision-making power and time efficiency.

@@

In 1939, Dirac proposed his notation~\cite{dirac1939new} for quantum mechanics, which is designed to represent linear algebra formulae in a compact and convenient way.
For instance, $a\ket{\psi} + b\ket{\phi}$ indicates the addition of two vectors, i.e., the superposition of two states $\ket{\psi}$ and $\ket{\phi}$.
Dirac notation is now widely accepted as the working language in quantum computation and quantum information. The reasonings of Dirac notation play a fundamental role in research and application, just like boolean and integer logic to classical computer science. 
For example, formalization works of quantum algorithms and quantum programming languages involves plenty of equational proofs of Dirac notation, which are critical, repeating but time-intensive jobs.
Dirac notation are also used in quantum programming languages to define the program states, operations and assertions. In order to automate the verification procedure, we need to simplify and check the equivalence of pre-conditions. 
However, unlike the existence of SAT and SMT solver, we are still in need of a working Dirac notation solver, and this absence has become an obstacle of many areas.

Recently, Xu et al.~\cite{diracdec} proposed a theory to decide the equivalence of Dirac notation, as well as a prototype implementation in Mathematica called DiracDec.
They proved that the equivalence of basic Dirac notation are decidable.
Their algorithm sticks with a pure term rewriting system, and is proved to be confluent and terminating.
Even though, there is still a gap between DiracDec and a practical solver for Dirac notation equivalence.

One problem is the algorithm efficiency for equivalence out of the scope of rewritings. DiracDec decides the whole equational theory by rewriting modulo $E$, where $E$ is a set of equational axioms that cannot be decided by normalization in rewritings, e.g. the associativity and commutativity (AC) of some function symbols. DiracDec uses a direct by inefficient algorithm to decide $E$, which search through all possible permutations and has factorial complexity.
This is witnessed in their evaluation: DiracDec is not good at dealing with ``computational examples'', which has many AC symbols and are therefore time consuming.

Usability is another problem.
DiracDec does not support labelled Dirac notation, which use registers to denote the subsystem, and express the state locally.
The typing of DiracDec only deals with Dirac notation, which is not sufficient for a working scenario, where we want to have defined symbols and a context for the typing assumptions of variables.
Moreover, to avoid type checking during term rewriting, DiracDec separate the same symbol (e.g. multiplication) for different types to disambiguate, which makes the system bloated.
Also, it is hard to integrate the Mathematica implementation as a solvers into other projects.

The system design involves a trade-off between simplicity and efficiency. The simplicity of term rewriting in DiracDec allows them to reach satisfying theoretical results, but it is also a constraint for optimization.
Based on DiracDec, this work aims at building a practical solver. It transforms the term rewriting system into a hybrid algorithm, and overcomes the problems mentioned above. Our main technical contributions are:
\begin{itemize}
    \item An efficient algorithm to decide the equational theories in $E$. The core idea is to perform equivalence checking through normalization, with the normal form for $E$ being obtained by sorting.
    \item The support for constant register labels. The equivalence decision of labelled Dirac notation is reduced to the unlabelled situation.
    \item A more usable \CC\ solver Dirace with abstract language and typing. We also support to define symbols (e.g., transpose and trace) using the functions syntax.
\end{itemize}

As an evaluation, we tested Dirace on the benchmark of DiracDec and new examples for labelled Dirac notation. 
The result shows significant improvement in decidability and efficiency over DiracDec.
% Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.

% \yx{mention the dilemma of simplicity and efficiency}

\section{Motivation and Preliminary}
One interesting fact about the quantum world is that for two maximally entangled states, applying one operator $M$ on one subsystem is equivalent to applying $M^T$, the transpose of $M$, on the other. This property holds no matter how far the two systems are separated, and is expressed as an equation of Dirac notation.
\begin{example}
    \label{ex: motivating}
    Let $q$ and $r$ be two quantum systems on $\mathcal{H}_T$ space. Let $M$ be a quantum operation on $\mathcal{H}_T$,
    and $\ket{\Phi} = \sum_{i \in T} \ket{i}\otimes \ket{i}$ be the maximally entangled state, then we have
    \[
    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
    \]
\end{example}
Here $q$ and $r$ are called labels, denoting the system for Dirac notation.
To automate reasoning about such equations, we need to formalize the language and develop a proof system.
% The following paragraphs briefly introduce the Dirac notation and the universal algebra framework we use.

\subsubsection{Dirac notation}
Quantum states live in complex Hilbert spaces. We use vectors in the space to descirbe pure quantum states, and operations are described by linear transformations. 
It uses the ket $\ket{i}$ and the bra $\bra{i}$ to indicate bases of the space and the dual space. Together with other variable symbols, they are composed with each other in sequence, and the composition will be interpreted into different operations, depending on the type of operands. For example, $\braket{i|j}$ represents the inner product of $\bra{i}$ and $\ket{j}$, which is a scalar, while $\ket{i}\bra{j}$ represents the outer product, resulting in an operator. 
Dirac notation also use $\otimes$ to indicate the vectors and operators in the tensor product space.

Dirac notation further enjoys the property that the interpretation is independent on the order of composition, thus parentheses can be omitted. For example, the formula \(\bra{i}\ket{\phi}\bra{\psi}\ket{j}\) can be understood as
\[
    \braket{i|\phi}\braket{\psi|j} = \bra{i} (\ket{\phi}\bra{\psi}) \ket{j},
\]
and they are equivalent for all variables.

% With the concrete basis, Dirac notation can be interpreted as matrices. For example, the ket of two basis of qubit states are column vectors $\ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix}$ and $\ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix}$, and the corresponding bras are row vectors in the dual space $\bra{0} = \begin{bmatrix} 1 & 0 \end{bmatrix}$ and $\bra{1} = \begin{bmatrix} 0 & 1 \end{bmatrix}$.
% The calculation between Dirac notation then corresponds to matrix operatoins. For example, inner product $\braket{0| 1} = \begin{bmatrix} 0 & 1 \end{bmatrix} \begin{bmatrix} 1 \\ 0 \end{bmatrix} = 0$, and tensor product $\ket{0} \otimes \ket{1} = \ket{(0, 1)} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \otimes \begin{bmatrix} 0 \\ 1 \end{bmatrix} = \begin{bmatrix} 0 \\ 1 \\ 0 \\ 0 \end{bmatrix}$.

In practice, Dirac notation are combined with other syntax, such as the big operator sum $\sum_{i \in S} A$, for better expressiveness. We also use labelled Dirac notation like $\ket{i}_q \otimes \ket{i}_r$ to denote the quantum system in consideration, where Dirac notation are subscripted with registers.

\subsubsection{Universal algebra}
We use universal algebra and equational logic to formally represent Dirac notation and the reasoning procedure.
A universal algebra defines a signature of function symbols.
Terms in the algebra are constructed by constants, variables or function application on terms. 
Other basic concepts like substitution of variables or pattern matching are also defined.
In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
Reasonings of Dirac notation terms follow the equational logic. It defines an equaitonal relation compatible with substitution and term construction, which essentially formalizes the common sense of equivalence in algebra.

\section{Language, Typing and Semantics}
The first step is to formally pin down the language for Dirac notation.
As a reference, the language in DiracDec has a concrete design, meaning that the same syntax for different types corresponds to different symbols.
Our intention is to move from the concrete design to a more abstract one. This is closer to the casual Dirac notation we use, and simplifies the term rewriting system.

For this purpose, our language involves three layers: the index, the type and the term.
Terms describe the concrete instances like ket, bra and operator, which will be checked and typed.
Index represents classical datatypes, and they appear in type expressions to distinguish between different Hilbert spaces and sets.
\begin{definition}[index syntax]
    The syntax for type indices is
    \begin{align*}
        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
    \end{align*}
\end{definition}
Here $x$ is a variable. $\sigma_1 \times \sigma_2$ is the product type for tensor product space or Cartesian product sets.

\begin{definition}[type syntax]
    The syntax for Dirac notation types is
    \begin{align*}
        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma). \\
    \end{align*}
\end{definition}
$\Basis(\sigma)$ is the type for basis in index $\sigma$. $\SType$ indicates scalars, $\KType(\sigma)$ and $\BType(\sigma)$ indicates ket and bra terms in Hilbert space $\sigma$, and $\OType(\sigma_1, \sigma_2)$ indicates linear operators with Hilbert space $\sigma_2$ as domain and $\sigma_1$ as codomain.
$\mathsf{Set}(\sigma)$ is the type of subsets in $\sigma$ index, and is used to denote the values of bound variables in summation.
The remaining two generators are types for functions. $T_1 \to T_2$ denotes normal functions, which accepts a $T_1$ type argument and results in a $T_2$ type term. $\forall x. T$ denotes index functions, which accepts an index argument $x$ and results in a $T$ type term, where $T$ can depend on $x$.
Index functions are necessary for defining transformations that are polymorphic on Hilbert spaces.

\begin{definition}[term syntax]
    The syntax for Dirac notation terms is
    \begin{align*}
        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
    \end{align*}
\end{definition}
The generators are explained in order.
$\lambda x : T.e$ is the abstraction for normal functions, and $\mu x.e$ is the abstraction for index functions.
$e_1\ e_2$ is function application. 
% $e_1 \circ e_2$ is the composition operator of Dirac notation.
% $\hat{0}$ and $\hat{1}$ are constant basis of $\Basis(\mathsf{bool})$ type.
$(e_1, e_2)$ is the basis pair for product types. $0$, $1$, $\ADDS$, $e_1 \times \cdots \times e_n$ and $e^*$ are symbols for scalars. 
The next line includes constant symbols for ket, bra and operator.
$e^\dagger$ denotes the conjugate transpose of $e$. $e_1.e_2$ denotes scaling the term $e_2$ with scalar $e_1$.
$\mathbf{U}(e)$ is the universal set with index $e$. $e_1 \star e_2$ is the Cartesian product. $\sum_{e_1} e_2$ is the big operator sum, which is modelled by folding the function $e_2$ over value sets $e_1$. Usually, the sum body is specified by an abstraction. Therefore we use the notation $\sum_{x \in s} X$ to denote $\sum_{s} \lambda x : T . X$ as well.
Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. 
% They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator. We encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.

Compared to DiracDec, the addition, adjoint, scaling and tensor product symbols are merged together.
They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.
The multiplications are still separated because their properties do not have to much overlap.

@@

Typing is the procedure to classify the term through a proof system in a typing context. 
Our context consists of the environment $E$ for definitions and assumptions, and the context $\Gamma$ for bound variables. 
$E$ and $\Gamma$ are sequences of assumptions $x : T$ or definitions $x := t : T$.
\begin{definition}[environment and context]
    \begin{align*}
        E ::=\ & [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
        \Gamma ::=\ & [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
    \end{align*}
\end{definition}
Definitions refer to the symbols that can be unfolded. Definition symbols usually have abstract meanings, and here we use them to encode more operations in Dirac notation, such as transpose and trace.
Assumptions are the declarations of the type for variables. In the scenario of equational proofs, variable assumptions are universally quantified implicitly.

The type checking of our language involves maintaining a well-formed environment and context $E[\Gamma]$.
We say an expression $t$ has type $X$ in context $E[\Gamma]$, if the typing judgement $E[\Gamma] \vdash t : X$ can be proved through the typing rules in~\Cref{sec: full typing rules}. Here we present and explain the rules selectively. Firstly, well-formed contexts $\WF(E)[\Gamma]$ are built in the incremental way, e.g.:
\[
    \frac{}{\WF([])[]}
    \qquad
    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
    \qquad
    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x:=t:T)[]}.
\]
Starting from an empty context, we can assume new index symbols, and assume or define symbols with checked types. Based on the well-formed context, typing judgements can be proved by information from $E[\Gamma]$, or built inductively. In the following rules, for example, the condition $x : \Index \in E[\Gamma]$ is true if $E$ or $\Gamma$ has the assumption in their sequences. $\KType(\sigma)$ will be a valid type for kets, if the argument $\sigma$ is typed as the index.
\begin{gather*}
    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
    \qquad
    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
\end{gather*}

The Dirac notation will then be typed accordingly. For example, the ket syntax $\ket{t}$ has type $\KType(\sigma)$, if $t$ is typed as a basis term of index $\sigma$. Also, the inner product of a bra and a ket with the same type index $\sigma$ is typed as the scalar. This corresponds to the constraint of inner products that vectors should be in the same Hilbert space.
\begin{gather*}
    \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)}
    \qquad
    \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
\end{gather*}

The typing for functions and applications follow the common practice. 
Take the index function $\mu x. t$ as an example: here $x$ is a bound variable typed as $\Index$, and the type $U\{x/u\}$ of application $(t\ u)$ is obtained by replacing $x$ with the index instance $u$.
\begin{gather*}
    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
    \qquad
    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} \\
    \\
    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
    \qquad
    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
\end{gather*}

Lastly, the big operator sum is modelled by folding a function on a set, therefore the typing rule is as follows:
\[
    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
\]

% And lastly we have the typing rules for composition $x \circ y$. Similar to casual Dirac notation, the typing of composition depends on the types of operands.
% \begin{gather*}
%     \frac{E[\Gamma] \vdash x : \SType \qquad E[\Gamma] \vdash y : \KType(\sigma)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
%     \qquad
%     \frac{E[\Gamma] \vdash x : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash y : \KType(\tau)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
% \end{gather*}

\begin{lemma}
    The typing of expressions are decidable and unique.
\end{lemma}
\begin{proof}
    The type can be calculated in the recursive manner. There is at most one typing rule for the same function symbol and argument types, which leads to the uniqueness of typing.
\end{proof}

@@

% Semantics define the meanings of expressions, and the goal of the desirable algorithm is to decide whether
% two expressions have the equivalent semantics. We can give the semantics by axiomatic equations and the denotational interpretations.

To give meanings to the syntax, we assign the semantics in the denotational way, which is a mapping from syntax to set-theoretical objects.

\subsubsection{Denotational semantics} 
The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping $v$ from variables to their values. 
The semantics of $e$ with valuation $v$ is written as $\sem{e}_v$, and two expressions $e_1$ and $e_2$ are equivalence if they have equal semantics in the mathematical sense for all valuations, written as $\sem{e_1} = \sem{e_2}$.

>>>>

%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Dirac notation is a fundamental language in quantum computation and quantum information. Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate equational reasoning with Dirac notation, a critical yet time-intensive task. Building upon DiracDec, this work aims to develop a solver optimized for practical applications. Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labeled Dirac notation. Our solver, Dirace, is implemented in \CC\ with a Mathematica backend, demonstrating significant improvements in decision-making power and time efficiency.

@@

In 1939, Dirac introduced his notation for quantum mechanics~\cite{dirac1939new}, designed to represent linear algebraic formulas in a compact and convenient form. For example, the expression \( a\ket{\psi} + b\ket{\phi} \) represents the superposition of two quantum states, \( \ket{\psi} \) and \( \ket{\phi} \). Today, Dirac notation is widely accepted as the standard language in quantum computation and quantum information. Its reasoning forms the foundation of research and applications, much like boolean and integer logic do in classical computer science.

In quantum algorithm formalizations and quantum programming languages, Dirac notation is frequently used in equational proofs, which are critical, repetitive, and often time-intensive. These notations also play a key role in defining program states, operations, and assertions in quantum programming languages. To automate the verification of these programs, we need tools that can simplify and check the equivalence of preconditions. However, unlike the well-established SAT and SMT solvers for classical logic, a practical solver for Dirac notation equivalence remains an unmet need, creating a barrier for progress in several fields.

Recently, Xu et al.~\cite{diracdec} proposed a theory for deciding the equivalence of Dirac notation, alongside a prototype implementation in Mathematica called DiracDec. They demonstrated that the equivalence of basic Dirac notation is decidable. Their algorithm, based on a pure term rewriting system, has been proven to be confluent and terminating. Despite this, there remains a gap between DiracDec and a practical solver for Dirac notation equivalence.

One challenge is the efficiency of the algorithm when dealing with equivalences beyond the scope of term rewriting. DiracDec decides the entire equational theory by rewriting modulo \( E \), where \( E \) represents a set of axioms that cannot be decided by normalization alone, such as the associativity and commutativity (AC) of certain function symbols. DiracDec uses a direct but inefficient algorithm to decide these axioms, which searches through all possible permutations and exhibits factorial complexity. This inefficiency becomes particularly evident in "computational examples" containing many AC symbols, which are time-consuming to process.

Usability is another area where DiracDec falls short. It does not support labelled Dirac notation, which uses registers to denote subsystems and express states locally. Additionally, DiracDec's typing system only does not provide context for variable typing assumptions or the definition of symbols, which are required in practical scenarios. To avoid type checking during term rewriting, DiracDec separates symbols (e.g., multiplication) for different types, leading to unnecessary complexity. Moreover, integrating the Mathematica implementation into other projects as a solver is challenging.

The system design of DiracDec reflects a trade-off between simplicity and efficiency. While the simplicity of term rewriting allows for strong theoretical results, it limits optimization opportunities. Building on DiracDec, this work aims to develop a practical solver. We transform the term rewriting system into a hybrid algorithm, overcoming the challenges mentioned above. Our main technical contributions are:
\begin{itemize}
    \item An efficient algorithm for deciding the equational theories in \( E \), based on equivalence checking through normalization, with the normal form for \( E \) being obtained via sorting.
    \item Support for constant register labels, and reducing the equivalence decision of labelled Dirac notation to the unlabelled case.
    \item A more user-friendly \CC\ solver, Dirace, featuring an abstract language and typing system. We also support the definition of symbols (e.g., transpose and trace) using function syntax.
\end{itemize}

We evaluated Dirace against the DiracDec benchmark and new examples involving labelled Dirac notation. The results show significant improvements in both decidability and efficiency compared to DiracDec.

% Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.

\section{Motivation and Preliminary}

An interesting property of quantum mechanics is that for two maximally entangled states, applying a quantum operator \( M \) to one subsystem is equivalent to applying \( M^T \) (the transpose of \( M \)) to the other subsystem. This relationship holds regardless of the spatial separation between the two systems, and it can be expressed as an equation in Dirac notation.
\begin{example}
    \label{ex: motivating}
    Let \( q \) and \( r \) represent two quantum systems in the Hilbert space \( \mathcal{H}_T \). Let \( M \) be a quantum operation acting on \( \mathcal{H}_T \), and let \( \ket{\Phi} = \sum_{i \in T} \ket{i} \otimes \ket{i} \) be the maximally entangled state. Then, we have the following equation:
    \[
    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
    \]
\end{example}

In this equation, \( q \) and \( r \) are labels denoting the respective subsystems in the Dirac notation. To automate reasoning about such equations, we must formalize the language and develop a proof system to handle it.

\subsubsection{Dirac Notation}

Quantum states are represented as vectors in complex Hilbert spaces, and operations on these states are described by linear transformations. Dirac notation uses the ket \( \ket{i} \) and the bra \( \bra{i} \) to denote basis vectors in a Hilbert space and its dual space, respectively. These symbols can be composed together in various ways to form more complex expressions. The meaning of these compositions depends on the types of the operands involved.
For example, the inner product \( \braket{i|j} \) represents the scalar result of the dot product between \( \bra{i} \) and \( \ket{j} \), while the outer product \( \ket{i}\bra{j} \) represents an operator. Additionally, Dirac notation uses the tensor product symbol \( \otimes \) to describe the combined space of multiple quantum systems.

One of the key features of Dirac notation is its order-independent interpretation. This means that the composition of terms can be written without parentheses, as the interpretation is unaffected by the order of multiplication. For instance, the formula \( \bra{i}\ket{\phi}\bra{\psi}\ket{j} \) can be understood as:
\[
    \braket{i|\phi}\braket{\psi|j} = \bra{i} (\ket{\phi}\bra{\psi}) \ket{j},
\]
and these expressions are equivalent for all variables involved.

In practice, Dirac notation is often combined with other syntactic elements, such as the summation symbol \( \sum_{i \in S} A \), to enhance expressiveness. Furthermore, labelled Dirac notation, such as \( \ket{i}_q \otimes \ket{i}_r \), is used to denote quantum systems in consideration, where subscripts (registers) are added to distinguish different subsystems.

\subsubsection{Universal Algebra}
We use universal algebra and equational logic to formally represent Dirac notation and the reasoning procedure. A universal algebra defines a signature of function symbols, with terms constructed from constants, variables, and function applications. 
Other basic concepts like substitution of variables or pattern matching are also defined.
In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
The reasoning process is guided by equational logic, which defines an equivalence relation that is compatible with substitution and term construction. This relation formalizes the intuitive concept of equivalence in algebra.

\section{Language, Typing and Semantics}
The first step is to formalize the language for Dirac notation. In DiracDec, the language has a concrete design, where the same syntax for different types corresponds to distinct symbols. Our goal is to transition from this concrete design to a more abstract one, which aligns more closely with the conventional Dirac notation we use and simplifies the term rewriting system.

To achieve this, we organize our language into three layers: the index, the type, and the term. Terms represent concrete instances such as kets, bras, and operators, which will be typed and checked. The index represents classical data types and appears in type expressions to differentiate between various Hilbert spaces and sets.

\begin{definition}[Index Syntax]
    The syntax for type indices is:
    \begin{align*}
        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
    \end{align*}
\end{definition}
Here, \( x \) is a variable, and \( \sigma_1 \times \sigma_2 \) represents the product type for tensor product spaces or Cartesian product sets.

\begin{definition}[Type Syntax]
    The syntax for Dirac notation types is:
    \begin{align*}
        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma).
    \end{align*}
\end{definition}
\( \Basis(\sigma) \) denotes the type for basis elements in the index \( \sigma \).
\( \SType \) represents scalars, while \( \KType(\sigma) \) and \( \BType(\sigma) \) refer to ket and bra types in the Hilbert space \( \sigma \), respectively.
\( \OType(\sigma_1, \sigma_2) \) represents linear operators with \( \sigma_2 \) as the domain and \( \sigma_1 \) as the codomain.
\( \SET(\sigma) \) refers to the type of subsets of \( \sigma \), used to denote the values of bound variables in summations.
The remaining two constructs define function types: \( T_1 \to T_2 \) represents normal functions that take a \( T_1 \)-type argument and return a \( T_2 \)-type term, while \( \forall x. T \) represents index functions that take an index argument \( x \) and produce a \( T \)-type term, which may depend on \( x \).
Index functions are essential for defining polymorphic transformations across Hilbert spaces.

\begin{definition}[Term Syntax]
    The syntax for Dirac notation terms is:
    \begin{align*}
        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n) \mid e_1 \times \cdots \times e_n \mid e^* \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
    \end{align*}
\end{definition}
The terms above are explained in order:
\( \lambda x : T.e \) represents the abstraction for normal functions, and \( \mu x.e \) represents the abstraction for index functions.
\( e_1\ e_2 \) denotes function application.
\( (e_1, e_2) \) is the basis pair for product types.
The constants \( 0 \), \( 1 \), \( \ADDS \), \( e_1 \times \cdots \times e_n \), and \( e^* \) are symbols for scalars.
The next line includes constant symbols for kets, bras, and operators.
\( \ket{e} \) is a ket, \( \bra{t} \) is a bra, and \( e^\dagger \) denotes the conjugate transpose of \( e \).
\( e_1.e_2 \) represents scaling the term \( e_2 \) by scalar \( e_1 \).
\( \mathbf{U}(e) \) denotes the universal set with index \( e \).
\( e_1 \star e_2 \) represents the Cartesian product of \( e_1 \) and \( e_2 \).
\( \sum_{e_1} e_2 \) is the big operator sum, modeled by folding the function \( e_2 \) over the value set \( e_1 \). Typically, the sum's body is given by an abstraction. For convenience, we also use the notation \( \sum_{x \in s} X \) to represent \( \sum_{s} \lambda x : T . X \).
Additionally, \( \ADDS \) and \( \ADD \) are distinct AC symbols: \( \ADDS \) is used for scalar addition and \( \ADD \) for linear algebra addition.
There are five kinds of linear algebraic multiplications among kets, bras, and operators. These are encoded using five different symbols: \( \DOT \), \( \MULK \), \( \MULB \), \( \OUTER \), and \( \MULO \).

Compared to DiracDec, the symbols for addition, adjoint, scaling, and tensor products have been consolidated.
The multiplications are still kept separate due to their differing properties.
These operations are denoted as \( B \cdot K \), \( K_1 \cdot K_2 \), \( B_1 \cdot B_2 \), \( K \cdot B \), and \( O_1 \cdot O_2 \), respectively.

@@

The typing system is responsible for classifying terms within a proof system, using a context that specifies the types of variables and definitions. The context is divided into two components: the environment \( E \), which includes definitions and assumptions, and the context \( \Gamma \), which tracks bound variables. Both \( E \) and \( \Gamma \) consist of sequences of assumptions \( x : T \) or definitions \( x := t : T \).
\begin{definition}[Environment and Context]
    The syntax for \( E \) and \( \Gamma \) is:
    \begin{align*}
        E ::= &\ [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
        \Gamma ::= &\ [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
    \end{align*}
\end{definition}
Definitions refer to symbols that can be expanded or unfolded, and typically represent abstract concepts, such as transpose or trace in Dirac notation. Assumptions, on the other hand, define the types of variables, and variable assumptions are implicitly universally quantified in the case of equational proofs.

Type checking in our language involves maintaining a well-formed environment and context \( E[\Gamma] \). We say an expression \( t \) has type \( X \) in context \( E[\Gamma] \) if the typing judgment \( E[\Gamma] \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}. The well-formedness of the context \( \WF(E)[\Gamma] \) is built incrementally:
\[
    \frac{}{\WF([])[]}
    \qquad
    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
    \qquad
    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x := t:T)[]}.
\]
Starting with an empty context, new index symbols can be introduced, and symbols with checked types can be defined. Typing judgments are then made inductively using the information from \( E[\Gamma] \). For instance, the rule \( x : \Index \in E[\Gamma] \) holds if \( E \) or \( \Gamma \) contains an assumption for \( x \).
And \(\KType(\sigma)\) is a valid type for kets, if the argument $\sigma$ is typed as an index.
\begin{gather*}
    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
    \qquad
    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
\end{gather*}

Terms are then typed accordingly.
For example, the ket \( \ket{t} \) will have the type \( \KType(\sigma) \) if \( t \) is a basis term of index \( \sigma \). Similarly, the inner product between a bra and a ket of the same index \( \sigma \) is typed as a scalar. It corresponds to the constraint of inner product that vectors should be from the same Hilbert space.
\begin{gather*}
    \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)}
    \qquad
    \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
\end{gather*}

The typing for functions and applications follows the standard practice. For example, the index function \( \mu x. t \) is typed with \( x \) as a bound variable of type \( \Index \), and the application \( (t\ u) \) has the type \( U\{x/u\} \), obtained by replacing \( x \) with the index instance \( u \).
\begin{gather*}
    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
    \qquad
    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} 
    \\[1em]
    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
    \qquad
    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
\end{gather*}

Finally, the big operator sum is modeled by folding a function over a set, with the typing rule as follows:
\[
    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
\]

\begin{lemma}
    The typing of expressions is both decidable and unique.
\end{lemma}

\begin{proof}
    The type of an expression can be determined recursively. For any given function symbol and argument types, there is at most one typing rule, ensuring the uniqueness of typing.
\end{proof}

@@

The semantics of a language define the meaning of its expressions. In this context, the objective of our algorithm is to determine whether two expressions are semantically equivalent. We define the semantics in a denotational manner, mapping syntax to set-theoretic objects.

\subsubsection{Denotational Semantics}
The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping \( v \), which assigns values to the variables in the expression. The semantics of an expression \( e \) with a given valuation \( v \) is denoted as \( \sem{e}_v \). Two expressions \( e_1 \) and \( e_2 \) are considered equivalent if their semantics are equal for all valuations, i.e., \( \sem{e_1}_v = \sem{e_2}_v \) for all \( v \).
